Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2016-01-13 notes/reviews/response_to_reviews.tex

\begin{document}

\maketitle

Thank you for taking the time to review our paper. Here is our responses to the points raised.

\meta{
\begin{itemize}
 \item different parameter settings for toy model
 \item change the rank after burn in to show not stuck
 \item plots
 \item collapsed Gibbs
 \item pseudo-code algorithm
 \item Givens
 \item Reviewer 2
 \item Reviewer 3
 \item Proof read, aiming to (i) clarify (2) correct typos (c) define everything when first used
\end{itemize}

@@

\begin{review}
It will be helpful to provide some relevant applications in Section I, explaining why a degenerate model will be more appropriate.
\end{review}

\begin{response}
We have modified the introduction to mention the example of tracking multiple points on a rigid body.
\end{response}

\begin{review}
There is no complexity analysis.
\end{review}

\begin{response}
There is a brief discussion of the computational complexity in the final section of the paper. There is not much to say -- it is of the same order for the full rank and degenerate models. We do not see much value in including a detailed operation count, since the dominating factor will be the difference in autocorrelation and the resulting difference in the necessary length of the chain, which will vary by model and data set.
\end{response}

@@

\section*{Response to Reviewer 2}

\begin{review}
This manuscript studies the problem of Gibbs sampling in the estimation of degenerate linear Gaussian state space model, where the covariance matrix Q of noise in state transition function is singular. This manuscript proposes to decompose matrix Q and convert the problem to a general case conditioned on the known rank $r$ of $Q$. In order to estimate rank $r$, reversible jump MCMC is employed. There are several concerns as follows:

1.  In the manuscript, the authors assume a linear model in which the observation is only determined by the current state. However, in practice, sophisticated models where the observation is also related to the previous states and inputs are more commonly used. So this may limit the applicability of the proposed method.
\end{review}

\begin{response}
 There is a very large body of literature focussed on models where the observation depends only on the current latent state, including most of the papers we have cited. It is usually possible to convert such a non-Markovian model into a Markovian one by augmenting the latent state with additional components. We disagree that this should be considered a limitation of the proposed method.
\end{response}

@@

\begin{review}
4.  This manuscript uses Givens decomposition to factorize matrix $Q$, and in section III B, it states that matrix $U$ in Givens decomposition is uniquely determined. It should be explained that how to determine matrix $U$ and the necessity of Givens decomposition.
\end{review}

\begin{response}
 \meta{Add more discussion of the Givens decomposition and add algorithm for generating it.}
\end{response}

\begin{review}
5.   The first step of Gibbs sampling in linear state space model is the estimation of states $x$ from observation $y$. It should be clarified that whether the singularity of matrix $Q$ will have any effect on this estimation. 
\end{review}

\begin{response}
 The procedure for sampling the state trajectories is unchanged by the singularity of $Q$. We have noted this in the text and also added a short appendix discussing this point and providing the equations for the forward-filtering-backward-sampling algorithm.
\end{response}

@@

\begin{review}
6.  It should be explained that how the probabilities of $P(r \to r+1)$ and $P(r+1 \to r)$ in Eq.(59) are determined in simulations.
\end{review}

\begin{response}
 These two probabilities are design parameters. We have clarified this and added their values to the simulations section.
\end{response}

\begin{review}
7.  In simulations, the Metropolis-Hastings step should also be applied to the full rank linear model if not.
\end{review}

\begin{response}
This comment is not entirely clear to us. We could indeed apply the random walk MH steps to the full rank model, but what would be the point? These steps are essential only in the degenerate case to ensure the reducibility of the Markov chain. The mixing due to the MH steps is much slower than that due to the Gibbs steps, so including them in the full rank algorithm has a negligible effect on the autocorrelation.
\end{response}

@@

\begin{review}
8.  In eq. (24), function $exp()$ should be replaced by $exp(tr())$. In line 38 page 10, the first 'is' should be deleted.
\end{review}

\begin{response}
 We have corrected these two typos.
\end{response}

\begin{review}
I suggest the authors to address these problems and to submit it to IEEE Signal Processing Letters or do a resubmission.
\end{review}

\begin{response}
 We considered submitting this as a letters paper but we do not believe it meets, or could be made to meet, the requirements for either length or content.
\end{response}

### 2016-01-13 notes/paper-degenerate-linear-systems.tex
More corrections: ffbs, and other small bits

An appropriate Gibbs sampler can be constructed by sampling alternately from the following conditional posterior distributions,
%
\begin{IEEEeqnarray}{l}
 \postden(\ls{1:\timax}|\lgtm, \lgtv) \nonumber \\
 \postden(\lgtm, \lgtv| \ls{1:\timax}) \nonumber      .
\end{IEEEeqnarray}
%
The distribution of the sampled values will then converge to the target posterior distribution \cite{Roberts1994}. The state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be sampled using the forward-filtering-backward-sampling algorithm \cite{Chib1996,Wills2012}, which works by first running a Kalman filter forwards through the data, then running backwards in time sampling each state conditional on those in the future. The parameter conditional may also be sampled directly. This depends on the state sequence probability,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\lgtv) \propto \prod_{\ti=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \\
 & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber      .
\end{IEEEeqnarray}

@@

As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$. Nevertheless, we can sample from this distribution exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions. These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.

@@

@@

The computational complexity of the algorithm is $\bigo{\lsd^3 \timax}$, the same as the standard Gibbs sampler for the full rank model. For moderate to large $\timax$, the limiting factor is the forward-filtering-backward-sampling steps used to draw from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$.

@@

>>>>

An appropriate Gibbs sampler can be constructed by sampling alternately from the following conditional posterior distributions,
%
\begin{IEEEeqnarray}{l}
 \postden(\ls{1:\timax}|\lgtm, \lgtv) \nonumber \\
 \postden(\lgtm, \lgtv| \ls{1:\timax}) \nonumber      .
\end{IEEEeqnarray}
%
The distribution of the sampled values will then converge to the target posterior distribution \cite{Roberts1994}. The state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be sampled using the forward-filtering-backward-sampling algorithm \cite{Chib1996,Wills2012}, which works by first running a Kalman filter forwards through the data, then running backwards in time sampling each state conditional on those in the future. See appendix~\ref{app:state-sampling}. The parameter conditional may also be sampled directly. This depends on the state sequence probability,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\lgtv) \propto \prod_{\ti=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \\
 & \propto & \determ{\lgtv}^{-\frac{\timax-1}{2}} \exp\left( -\half \sum_{\ti=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \lgtv\inv (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber      .
\end{IEEEeqnarray}

@@

As discussed, standard Gibbs sampling where we alternately sample from the state and parameter conditionals does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$. Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions. These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.

Note that the forward-filtering-backward-sampling procedure can be used as before to draw from the state conditional. See appendix~\ref{app:state-sampling}. The parameter conditional, on the other hand, will require some modifications.

@@

%TODO: move this
\subsubsection{Forward Filtering Backward Sampling}

The forward-filtering-backward-sampling procedure for drawing from $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ can be applied to a degenerate model without modification. The Kalman filter stage is

@@

@@

\section{Sampling the State Trajectory} \label{app:state-sampling}

The Kalman filter recursively calculates the state predictive and filtering distributions,
%
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ob{1:\ti-1}, \lgtm, \lgtv) &=& \normalden{\ls{\ti}}{\lspmn{\ti}}{\lspvr{\ti}} \\
 \den(\ls{\ti}|\ob{1:\ti}, \lgtm, \lgtv) &=& \normalden{\ls{\ti}}{\lsfmn{\ti}}{\lsfvr{\ti} \vphantom{\lspvr{\ti}} }     .
\end{IEEEeqnarray}
%
The recursion equations for the means and variances are well known. See e.g. \cite{Sarkka2013}.

The forward-filtering-backward-sampling procedure draws a sample from $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$. This is achieved by first sampling from the final filtering distribution $\den(\ls{\timax}|\ob{1:\timax})$ and then for $\ti=\timax-1 , \dots , 1$ sampling from $\den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv)$. Each of these factors is also a Gaussian,
%
\begin{IEEEeqnarray}{l}
 \den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv) \nonumber \\
 \quad = \normalden{\ls{\ti}}{ \lsfmn{\ti} + K_t (\ls{\ti+1}-\lgtm\lsfmn{\ti}) }{ \lspvr{\ti} - K_t \lgtm \lsfvr{\ti} }      ,
\end{IEEEeqnarray}
%
where $K_t = \lsfvr{\ti} \lgtm\tr (\lgtm \lsfvr{\ti} \lgtm\tr + \lgtv)\inv$. The proof is straightforward using standard identities for the Gaussian distribution \cite{Sarkka2013}.

This procedure is still valid when $\lgtv$ is singular, provided the matrix $(\lgtm \lsfvr{\ti} \lgtm\tr + \lgtv)$ is invertible. A sufficient condition for this is that both the initial covariance $\lsfvr{1}$ and the transition matrix $\lgtm$ are both non-singular. With a matrix normal prior on $\lgtm$, the probability of it being singular is almost surely $0$.

To derive the expression for $\den(\ls{\ti}|\ls{\ti+1}, \ob{1:\ti}, \lgtm, \lgtv)$ in the case where the transition covariance is singular, replace $\lgtv$ with $\lgtv + h I$ where $I$ is the identity matrix and $h$ a non-negative real scalar such that a transition density can be written down. Then the same algebraic steps can be followed as in the non-singular case. At the end, set $h=0$ and obtain the same formula.
